---
title: "p8105_hw2_ryd2106"
author: "Risa Dickerman"
date: "2026-09-30"
output: github_document
---
Library load in
```{r setup, include=FALSE}
library(tidyverse)
```

# Problem 1

Load in and transform data
```{r}
#Read in dataset
train_data = read_csv("./data_hw/NYC_Transit_Data.csv",
                na = c(".", "NA", ""),
            col_types = cols(
              Route8 = col_character(),
              Route9 = col_character(),
              Route10 = col_character(),
              Route11 = col_character())) |> 
#clean the data set
  janitor::clean_names() |> 
#vector selection
  select(line:route11,vending,entry,entrance_type,ada) |> 
#transforming entry vector entries from Yes/No to True/False  
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE)) |> 
#Convert the entry vector from character to logical
  mutate(
    entry = as.logical(entry)
  )
head(train_data)
#Quiet uneccesary warning message
show_col_types = FALSE
  
```

The resultant dataset contains 19 different variables. Of which, the following 
are character types: line, station_name, each of the 11 routes, vending, 
and entrance_type. The numeric vectors are station_latitude, and
station_longitude. The logical vectors are: entry, and ada. So far, I have
tidied the dataset with the janitor function. I then, converted the Yes/No 
character values in the entry vector to TRUE/FALSE logical values. To ensure all
routes were character values and not partially numeric, I converted their value
types to character ones.Finally, I mutated this entry vector type from character 
to logical. The final data set is a (1868 x 19) dataset. This dataset is not tidy 
since the route data is spread over multiple columns when it is technically a 
single variable. To fix this, the route data should be condensed into two 
variable columns. One stating the number of routes passing through the station, 
the other stating what these routes are. Additionally, the station name is 
separated into two columns and should be combined into one. Finally there are 
many repeats of the same entry. To fix this the distinct function should be 
applied to get only one true entry value.

```{r}
#Numer of unique stations
combined_stations = unite(train_data, station_united, line, station_name) 
distinct_stations = combined_stations |> 
  unique() 
#Number of ADA compiant stations
ada_distinct = distinct_stations |> 
  filter(ada == TRUE) |> 
  nrow()

#Portion of stations entrances / exists without vending allow entrance
entrance_vending = combined_stations |> 
  distinct(station_united, entry, vending) |> 
  mutate(
   vending = case_match( 
     vending,
      "YES" ~ 1,
      "NO"  ~ 0)) |> 
  mutate(
    entry = as.numeric(entry)) 
#Numeric sum of entry data
entry_count = combined_stations |>
  pull(entry) |> 
  sum(na.rm = TRUE) 
#Numeric sum of vending
vending_count = entrance_vending |>
  mutate(
    vending = case_match(
      vending,
      1 ~ 1,
      0 ~ 0,
      NaN ~ 0)) |> 
  pull(vending) |> 
  sum(na.rm = TRUE) 

portion_sum =  entry_count/ vending_count
  


```

There are `r nrow(distinct_stations)` distinct stations, of these `r ada_distinct`
are ADA compliant. The proportion of station entrances / exits without venting
is `r portion_sum`.


```{r}
rerouted_train = distinct_stations |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name")

atrain_stations = rerouted_train |> 
  filter(route_name == "A") |> 
  nrow()

```
There are `r atrain_stations` distinct stations that serve the A train.



